## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10
if (a <=5) {
    print(a)
}   else {
    print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5) {
    print("small")
}   else if (a > 5 & a <= 10) {
    print("medium")
}   else if (a > 10 | a == 7) {
    print("large")
}   else {
    print("huge")
}


```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a / 2 != 0) {
    if (a ==5){
        print("a is 5")
    }
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){
    # paste is concat function
    print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1,1)

#change 1st, 3rd, and 6th i to 6

for (i in c(1,3,6)){
    a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow =2, ncol =2)

for (row in 1:2){
    for (col in 1:2){
        mat[col,row] <-  0
        print(mat)
    }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10) {
    b <- b + 2 # add 2 to b
    print(b) # print b
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
#start with a vector temperatures

test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

c_to_f <- function(dat){
    temp_fahr <- dat*(9/5) + 32
    
    return(temp_fahr)
}


```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)

apply(test_mat, 2, sum)

```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <-  matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)

```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}

    # Provided url links for last two questions since instructions were a tad lengthy and not as straightforward as first 3 problems

    # CodeWars #1
    # Convert a number to a character

    number_to_character <- function(n) {
      return(as.character(n))
    }

    # Codewars #2
    # Find sum of all multiples of 3 or 5 BELOW the number passed into function
    # example; if 10 is value passed in it should have output of 23 (9,6,5,3)
    solution <- function(number){
      answer <- 0
      for (num in 1:number-1){
        if (num %% 3 == 0){
          answer = answer + num
        } else if (num %% 5 == 0)
          answer = answer + num
      }
      return(answer)
    }


    # CodeWars #3
    # Return the middle character of a given word. If word has even number of letters return the middle two characters 

    get_middle <- function(s){
      if (nchar(s) %% 2 ==0) {
        mid_2 <- substr(s, (nchar(s)/2), (nchar(s)/2 + 1))
        return(mid_2)
      } else if (nchar(s) %% 2 != 0) {
        mid <- substr(s, nchar(s)/2 + .5, nchar(s)/2 + .5)
        return(mid)
      }
    }


    # CodeWars #4
    # Take a Ten Minute Walk (6 kyu)
    # https://www.codewars.com/kata/54da539698b8a2ad76000228/train/r

    isValidWalk <- function(walk){
      if (length(walk) != 10) {
        return(FALSE)
      } else {
        north <- which("n" == walk)
        south <- which("s" == walk)
        west <- which("w" == walk)
        east <- which("e" == walk)
        if (length(north) == length(south) & length(west) == length(east)){
          return(TRUE)
        } else {
          return(FALSE)
        }
      }
    }

    # CodeWars #5
    # Counting the # of Duplicates (6kyu)
    # https://www.codewars.com/kata/54bf1c2cd5b56cc47f0007a1/train/r

    library(stringr)

    duplicate_count <- function(text){
      count <- 0
      already_checked <- c()
      for (char in 1:nchar(text)){
        dup <- tolower(substr(text, char, char))
        dup_check <- str_count(tolower(text), dup)
        if (dup_check > 1 & dup %in% already_checked == FALSE) {
          count <- count + 1
        }
        already_checked <- c(already_checked, dup)
      }
      return(count)
    }


    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}

    fizzybuzzy <- function(number){
      if (number %% 3 == 0 & number %% 5 == 0){
          return("fizzbuzz")
    } else if (number %% 5 == 0) {
          return("buzz")
    } else if (number %% 3 == 0) {
          return("fizz")
    } else {
          return("Not a multiple of 3 or 5")
    }
    }

    fizzybuzzy(13)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.


    marioStairs <- function(number){
        for (stair in 1:number){
            print(strrep('X', stair))
        }
    }

    marioStairs(9)

    #not sure if that's exactly what you were looking for but seemed to get job done
    ```
